Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories

### 2020-09-04 paper/main.tex
JRM final pass through main text, updated diff file

\begin{abstract}
The mental contexts in which we interpret experiences are often person-specific, even when the experiences themselves are shared. We developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences. We model experiences and memories as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration. Memory encoding can then be modeled as geometrically preserving or distorting the \textit{shape} of the original experience. We applied our approach to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging. Participants’ recountings all preserved coarse spatial properties (essential narrative elements), but not fine spatial scale (low-level) details, of the episode’s trajectory. We also identified networks of brain structures sensitive to these trajectory shapes. Our work provides insights into how our brains preserve and distort our ongoing experiences when we encode them into episodic memories.
\end{abstract}

@@

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway;][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

Comparing the overall shapes of the topic trajectories for the episode and participants' recalls reveals which aspects of the episode's essence were preserved (or discarded) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the \textit{precision} with which a participant recounts details about each event, and (2) the \textit{distinctiveness} of their recall for each event, relative to other events.  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list-learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode and/or how participants would later recount it.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually delineated) time segments spanning the roughly 50 minute video used in their study.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number of topics (typically one or two) tended to be ``active'' in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to) 10 sentences from their transcript as a document, and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared topic space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep[for review, see][]{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both \cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.005$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015$) numbers of recalled events.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (P17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across (Fisher $z$-transformed) correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\textit{Precision} is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in a participant's recounting.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  In other words, given that a recalled event best matches a particular episode event, more precisely recalled events overlap more strongly with the conceptual content of the original episode event.  When a given event is assigned a blend of several topics, as is often the case (Fig.~\ref{fig:model}), a high precision score requires recapitulating the relative topic proportions during recall.

@@

Figure \ref{fig:precision-detail} also illustrates the differences between high and low distinctiveness scores.  We extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) corresponding to the example episode event shown in Panel C (event 21).  We also extracted the annotations for all episode events whose content these participants' single recall events described .  We assigned each episode event a unique color (Fig.~\ref{fig:precision-detail}E), and colored each recall sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 21, whereas the least distinctive recall incorporates content from a much wider range of episode events.

@@

Visual inspection of the episode and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the episode (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in the reduced topic space, could the position of their \textit{next} recalled event be predicted reliably?  For each location in the reduced topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods} for additional details).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B reflect significant peaks at $p < 0.05$, corrected).  We observed that the locations traversed by nearly the entire episode trajectory exhibited such peaks.  In other words, participants' recalls exhibited similar trajectories to each other that also matched the trajectory of the original episode (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of HMM-identified recall events (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the episode.  Differences in the numbers of recall events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of disentangling classic ``proportion recalled'' measures (i.e., the proportion of episode events described in participants' recalls) from participants' abilities to recapitulate the episode's essence (i.e., the similarity between the shapes of the original episode trajectory and that defined by each participant's recounting of the episode).

In addition to enabling us to visualize the episode's high-level essence, describing the episode as a geometric trajectory also enables us to drill down to individual words and quantify how each word relates to the memorability of each event.  This provides another approach to examining participants' recall for low-level details beyond the precision and distinctiveness measures we defined above.  The results displayed in Figures \ref{fig:list-learning}C and \ref{fig:precision-detail}A suggest that certain events were remembered better than others.  Given this, we next asked asked whether the events were generally remembered precisely or imprecisely tended to reflect particular content.  Because our analysis framework projects the dynamic episode content and participants' recalls into a shared space, and because the dimensions of that space represent topics (which are, in turn, sets of weights over known words in the vocabulary), we are able to recover the weighted combination of words that make up any point (i.e., topic vector) in this space.  We first computed the average precision with which participants recalled each of the 30 episode events (Fig.~\ref{fig:topics}A; note that this result is analogous to a serial position curve created from our precision metric).  We then computed a weighted average of the topic vectors for each episode event, where the weights reflected how precisely each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by more precisely-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full episode, content that weighted heavily on topics and words central to the major foci of the episode (e.g., the names of the two main characters, ``Sherlock'' and ``John,'' and the address of a major recurring location, ``221B Baker Street'') were best remembered.  An analogous analysis revealed which themes were less-precisely remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to its average precision (Fig.~\ref{fig:topics}B, red box).  The least precisely remembered episode content reflected information that was extraneous to the episode's essence, such as the proper names of relatively minor characters (e.g., ``Mike,'' ``Molly,'' and ``Lestrade'') and locations (e.g., ``St. Bartholomew's Hospital'').

@@

A similar result emerged from assessing the topic vectors for individual episode and recall events (Fig.~\ref{fig:topics}C).  Here, for each of the three most and least precisely remembered episode events, we have constructed two wordles: one from the original episode event's topic vector (left) and a second from the average recall topic vector for that event (right).  The three most precisely remembered events (circled in green) correspond to scenes integral to the central plot-line: a mysterious figure spying on John in a phone booth; John meeting Sherlock at Baker St.\~to discuss the murders; and Sherlock laying a trap to catch the killer.  Meanwhile, the least precisely remembered events (circled in red) reflect scenes that comprise minor plot points: a video of singing cartoon characters that participants viewed in an introductory clip prior to the main episode; John asking Molly about Sherlock's habit of over-analyzing people; and Sherlock noticing evidence of Anderson's and Donovan's affair.

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories.  We next carried out a series of analyses aimed at understanding which brain structures might facilitate these preservations and transformations between the participants' shared experience of watching the episode and their subsequent memories of the episode.  In the first analysis, we sought to identify brain structures that were sensitive to the dynamic unfolding of the episode's content, as characterized by its topic trajectory.  We used a searchlight procedure to identify clusters of voxels whose activity patterns displayed a proximal temporal correlation structure (as participants watched the episode) matching that of the original episode's topic proportions (Fig.~\ref{fig:brainz}A; see \textit{Methods} for additional details).  In a second analysis, we sought to identify brain structures whose responses (during episode viewing) reflected how each participant would later structure their \textit{recounting} of the episode.  We used a searchlight procedure to identify clusters of voxels whose proximal temporal correlation matrices matched that of the topic proportions matrix for each participant's recall transcript (Figs.~\ref{fig:brainz}B; see \textit{Methods} for additional details).  To ensure our searchlight procedure identified regions \textit{specifically} sensitive to the temporal structure of the episode or recalls (i.e., rather than those with a temporal autocorrelation length similar to that of the episode and recalls), we performed a phase shift-based permutation correction (see \textit{Methods} for additional details). As shown in Figure~\ref{fig:brainz}C, the episode-driven searchlight analysis revealed a distributed network of regions that may play a role in processing information relevant to the narrative structure of the episode.  Similarly, the recall-driven searchlight analysis revealed a second network of regions (Fig.~\ref{fig:brainz}D) that may facilitate a person-specific transformation of one's experience into memory.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory effect analyses}~\citep[e.g.,][]{PallWagn02} to the domain of naturalistic experiences.

@@

The neural analyses we carried out (Fig.~\ref{fig:brainz}) also leveraged our geometric framework for characterizing the shapes of the episode and participants' recountings.  We identified one network of regions whose responses tracked with temporal correlations in the conceptual content of the episode (as quantified by topic models applied to a set of annotations about the episode).  This network included orbitofrontal cortex, ventromedial prefrontal cortex, striatum, among others.  As reviewed by \cite{RangRitc12}, several of these regions are members of the \textit{anterior temporal system}, which has been implicated in assessing and processing the familiarity of ongoing experiences, emotions, social cognition, and reward.  A second network we identified tracked with temporal correlations in the idiosyncratic conceptual content of participants' subsequent recountings of the episode.  This network included occipital cortex, extrastriate cortex, fusiform gyrus, and the precuneus.  Several of these regions are members of the \textit{posterior medial system}~\citep{RangRitc12}, which has been implicated in matching incoming cues about the current situation to internally maintained \textit{situation models} that specify the parameters and expectations inherent to the current situation~\citep[also see][]{ZackEtal07, ZwaaRadv98}.  Taken together, our results support the notion that these two (partially overlapping) networks work in coordination to make sense of our ongoing experiences, distort them in a way that links them with our prior knowledge and experiences, and encodes those distorted representations into memory for our later use.

>>>>

\begin{abstract}
The mental contexts in which we interpret experiences are often person-specific, even when the experiences themselves are shared. We developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences. We model experiences and memories as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration. Memory encoding can then be modeled as geometrically preserving or distorting the \textit{shape} of the original experience. We applied our approach to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging. Participants’ recountings all preserved coarse spatial properties (essential narrative elements), but not fine spatial scale (low-level) details, of the episode’s trajectory. We also identified networks of brain structures sensitive to these trajectory shapes. Our work provides insights into how we preserve and distort our ongoing experiences when we encode them into episodic memories.
\end{abstract}

@@

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway;][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect its low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

Comparing the overall shapes of the topic trajectories for the episode and participants' recalls reveals which aspects of the episode's essence were preserved (or lost) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the \textit{precision} with which a participant recounts details about each event, and (2) the \textit{distinctiveness} of their recall for each event, relative to other events.  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list-learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode and/or how participants would later recount it.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually delineated) time segments spanning the roughly 50 minute video used in their study.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number of topics (typically one or two) tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods}).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to) 10 sentences from their transcript as a document, and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content, quality, and quantity from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared topic space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep[for review, see][]{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both \cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.005$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015$) numbers of recalled events.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (P17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across the (Fisher $z$-transformed) correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\textit{Precision} is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in a participant's recounting.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  In other words, given that a recall event best matches a particular episode event, more precisely recalled events overlap more strongly with the conceptual content of the original episode event.  When a given event is assigned a blend of several topics, as is often the case (Fig.~\ref{fig:model}), a high precision score requires recapitulating the relative topic proportions during recall.

@@

Figure \ref{fig:precision-detail} also illustrates the differences between high and low distinctiveness scores.  We extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) corresponding to the example episode event shown in Panel C (event 21).  We also extracted the annotations for all episode events whose content these participants' single recall events described.  We assigned each episode event a unique color (Fig.~\ref{fig:precision-detail}E), and colored each recalled sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 21, whereas the least distinctive recall incorporates content from a much wider range of episode events.

@@

Visual inspection of the episode and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the episode (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in the reduced topic space, could the position of their \textit{next} recalled event be predicted reliably?  For each location in the reduced topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods}).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B reflect significant peaks at $p < 0.05$, corrected).  We observed that the locations traversed by nearly the entire episode trajectory exhibited such peaks.  In other words, participants' recalls exhibited similar trajectories to each other that also matched the trajectory of the original episode (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of HMM-identified recall events (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the episode.  Differences in the numbers of recall events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of disentangling classic ``proportion recalled'' measures (i.e., the proportion of episode events described in participants' recalls) from participants' abilities to recapitulate the episode's essence (i.e., the similarity between the shapes of the original episode trajectory and that defined by each participant's recounting of the episode).

In addition to enabling us to visualize the episode's high-level essence, describing the episode as a geometric trajectory also enables us to drill down to individual words and quantify how each word relates to the memorability of each event.  This provides another approach to examining participants' recall for low-level details beyond the precision and distinctiveness measures we defined above.  The results displayed in Figures \ref{fig:list-learning}C and \ref{fig:precision-detail}A suggest that certain events were remembered better than others.  Given this, we next asked asked whether the events that were generally remembered precisely or imprecisely tended to reflect particular content.  Because our analysis framework projects the dynamic episode content and participants' recalls into a shared space, and because the dimensions of that space represent topics (which are, in turn, sets of weights over known words in the vocabulary), we are able to recover the weighted combination of words that make up any point (i.e., topic vector) in this space.  We first computed the average precision with which participants recalled each of the 30 episode events (Fig.~\ref{fig:topics}A; note that this result is analogous to a serial position curve created from our precision metric).  We then computed a weighted average of the topic vectors for each episode event, where the weights reflected how precisely each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by more precisely-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full episode, content that weighted heavily on topics and words central to the major foci of the episode (e.g., the names of the two main characters, ``Sherlock'' and ``John,'' and the address of a major recurring location, ``221B Baker Street'') was best remembered.  An analogous analysis revealed which themes were less-precisely remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to its average precision (Fig.~\ref{fig:topics}B, red box).  The least precisely remembered episode content reflected information that was extraneous to the episode's essence, such as the proper names of relatively minor characters (e.g., ``Mike,'' ``Molly,'' and ``Lestrade'') and locations (e.g., ``St. Bartholomew's Hospital'').

@@

A similar result emerged from assessing the topic vectors for individual episode and recall events (Fig.~\ref{fig:topics}C).  Here, for each of the three most and least precisely remembered episode events, we have constructed two wordles: one from the original episode event's topic vector (left) and a second from the average recall topic vector for that event (right).  The three most precisely remembered events (circled in green) correspond to scenes integral to the central plot-line: a mysterious figure spying on John in a phone booth; John meeting Sherlock at Baker St.~to discuss the murders; and Sherlock laying a trap to catch the killer.  Meanwhile, the least precisely remembered events (circled in red) reflect scenes that comprise minor plot points: a video of singing cartoon characters that participants viewed in an introductory clip prior to the main episode; John asking Molly about Sherlock's habit of over-analyzing people; and Sherlock noticing evidence of Anderson's and Donovan's affair.

The results this far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories.  We next carried out a series of analyses aimed at understanding which brain structures might facilitate these preservations and transformations between the participants' shared experience of watching the episode and their subsequent memories of the episode.  In the first analysis, we sought to identify brain structures that were sensitive to the dynamic unfolding of the episode's content, as characterized by its topic trajectory.  We used a searchlight procedure to identify clusters of voxels whose activity patterns displayed a proximal temporal correlation structure (as participants watched the episode) matching that of the original episode's topic proportions (Fig.~\ref{fig:brainz}A; see \textit{Methods} for additional details).  In a second analysis, we sought to identify brain structures whose responses (during episode viewing) reflected how each participant would later structure their \textit{recounting} of the episode.  We used a searchlight procedure to identify clusters of voxels whose proximal temporal correlation matrices matched that of the topic proportions matrix for each participant's recall transcript (Figs.~\ref{fig:brainz}B; see \textit{Methods} for additional details).  To ensure our searchlight procedure identified regions \textit{specifically} sensitive to the temporal structure of the episode or recalls (i.e., rather than those with a temporal autocorrelation length similar to that of the episode and recalls), we performed a phase shift-based permutation correction (see \textit{Methods}). As shown in Figure~\ref{fig:brainz}C, the episode-driven searchlight analysis revealed a distributed network of regions that may play a role in processing information relevant to the narrative structure of the episode.  The recall-driven searchlight analysis revealed a second network of regions (Fig.~\ref{fig:brainz}D) that may facilitate a person-specific transformation of one's experience into memory.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory effect analyses}~\citep[e.g.,][]{PallWagn02} to the domain of naturalistic experiences.

@@

The neural analyses we carried out (Fig.~\ref{fig:brainz}) also leveraged our geometric framework for characterizing the shapes of the episode and participants' recountings.  We identified one network of regions whose responses tracked with temporal correlations in the conceptual content of the episode (as quantified by topic models applied to a set of annotations about the episode).  This network included orbitofrontal cortex, ventromedial prefrontal cortex, and striatum, among others.  As reviewed by \cite{RangRitc12}, several of these regions are members of the \textit{anterior temporal system}, which has been implicated in assessing and processing the familiarity of ongoing experiences, emotions, social cognition, and reward.  A second network we identified tracked with temporal correlations in the idiosyncratic conceptual content of participants' subsequent recountings of the episode.  This network included occipital cortex, extrastriate cortex, fusiform gyrus, and the precuneus.  Several of these regions are members of the \textit{posterior medial system}~\citep{RangRitc12}, which has been implicated in matching incoming cues about the current situation to internally maintained \textit{situation models} that specify the parameters and expectations inherent to the current situation~\citep[also see][]{ZackEtal07, ZwaaRadv98}.  Taken together, our results support the notion that these two (partially overlapping) networks work in coordination to make sense of our ongoing experiences, distort them in a way that links them with our prior knowledge and experiences, and encodes those distorted representations into memory for our later use.
